\documentclass[a4paper,UKenglish]{lipics}
%This is a template for producing LIPIcs articles. 
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
% for section-numbered lemmas etc., use "numberwithinsect"
 
\usepackage{microtype}%if unwanted, comment out or use option "draft"
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{fixltx2e}

\input{macros}

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{{An upper bound of 84 for Morpion Solitaire 5D}}%\footnote{This work was partially supported by someone.}}
\titlerunning{A Morpion 5D bound} %optional, in case that the title is too long; the running title should fit into the top page column

\author[1]{Henryk Michalewski}
\author[1]{Andrzej Nagórko}
\author[1]{Jakub Pawlewicz}
\affil[1]{Department of Mathematics, Informatics and Mechanics\\ University of Warsaw\\ \{H.Michalewski,A.Nagorko,J.Pawlewicz\}@mimuw.edu.pl} %\\ \texttt{open@dummyuni.org}}
\authorrunning{H. Michalewski, A. Nagórko and J. Pawlewicz} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

\Copyright{Henryk Michalewski, Andrzej Nagórko and Jakub Pawlewicz}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\subjclass{Dummy classification -- please refer to \url{http://www.acm.org/about/class/ccs98-html}}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
\keywords{Morpion, linear optmization, relaxation}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\serieslogo{}%please provide filename (without suffix)
\volumeinfo%(easychair interface)
  {Billy Editor and Bill Editors}% editors
  {2}% number of editors: 1, 2, ....
  {Conference title on which this volume is based on}% event
  {1}% volume
  {1}% issue
  {1}% starting page number
\EventShortName{}
\DOI{10.4230/LIPIcs.xxx.yyy.p}% to be completed by the volume editor
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\begin{abstract} 
The Morpion Solitaire is a paper-and-pencil single-player game played on a square grid with initial position consisting of $36$ dots.
In each move the player puts a dot on an unused grid position and draws a line that 
  consists of four consecutive segments passing through the dot.
 The goal is to find the longest possible sequence of moves.
There are two main variants of the game: 5T and 5D. 
They have different restrictions on how the moves may intersect.

Providing lower and upper bounds in Morpion Solitaire is a significant computational and mathematical challenge
  that led to a discovery of an important new algorithm.
 Best lower bounds for Morpion Solitaire were found in $2011$ by Rosin \cite{rosin} using a variant of the Monte Carlo method: 
 $178$ moves for the 5T variant, $82$ moves for the 5D variant.
 
% In the Morpion 5T variant an upper bound of $705$ was proved in $2006$ by Demaine at al. \cite{demaine}  using a method of potential. 
% A refinement of this approach was used in $2015$  in \cite{ijcai} to show a bound of $485$ - this method combined an isoperimetric inequality with linear programming. 

In the Morpion 5D variant 
Kawamura et al. proved in \cite{japonczycy} an upper bound of $121$ moves using a geometrical argument. 
We improve this bound to \therecord. 
This is done in two steps.
1) We state, using linear constraints, a geometric property that defines a class of graphs that includes all Morpion  positions - finding practically solvable linear constraints is a main conceptual advance of the present paper.
2) We solve the mixed integer problems and obtain 
an upper bound of \therecord for Morpion 5D. %The same method a) fully solves Morpion 5D with center symmetry (the optimal sequence is $68$ moves long) and b) gives an upper bound of $222$ for Morpion 5T with center symmetry. 
\end{abstract}

\input{introduction}
  
\input{geometry}

\input{linear}
  
\input{gemmating}

\input{upper_bounds}

\input{final}

\printbibliography 

\newpage
\appendix

\input{symmetry}
    
\end{document}

